R code from PRACTICE sections

### 2024-06-12 full_simulation.qmd
update rcode.qmd and full_simulation.qmd

---
title: "Full simulation"
---

::: {.callout-note}
This Notebook is used to run the full simulation for the case study discussed in the manuscript. The simulation takes several hours to run on a Mac Book Pro (13-inch, M1, 2020) with 16 GB memory.
:::

>>>>

---
title: "Run the full simulation for the case study example"
execute: 
  eval: false
---

::: {.callout-note}
This Notebook was used to run the full simulation for the case study discussed in the manuscript. The simulation takes more than 1 hour to run on a Mac Book Pro (13-inch, M1, 2020) with 16 GB memory.
:::

### 2024-06-12 rcode.qmd
update rcode.qmd and full_simulation.qmd

---
title: "R code from the practice sessions"
---

::: {.callout-note}
This Notebook contains only the R code from the practice sessions of the manuscript. The number of samples and repetitions for all simulations have been greatly reduced to ensure that the code runs quickly and with limited computational resources.
:::

@@

```{r}
simulate <- function(n_subjects = 100, n_items = 50,
  b_0 = 0.847, b_e = 1.350, b_a = -1.253, b_c = 2.603,
  b_ea = 0.790, b_ec = -1.393,
  sd_u0s = 0.5, sd_u0i = 0.5, ...){
  require(dplyr)
  require(faux)
  # simulate design
  dat <- add_random(subject = n_subjects, item = n_items) %>%
    add_between("subject", expert = c(1, 0), .prob = c(0.25, 0.75)) %>%
    mutate(advice_present = rbinom(n(), 1, prob = 2/3)) %>%
    mutate(advice_correct = if_else(advice_present == 1L, 
                                    rbinom(n(), 1L, prob = 0.8), 0L)) %>%
    # add random effects
    add_ranef("subject", u0s = sd_u0s) %>%
    add_ranef("item", u0i = sd_u0i) %>%
    # compute dependent variable
    mutate(linpred = b_0 + u0i + u0s +
        b_e * expert + b_a * advice_present + b_c * advice_correct +
        b_ea * expert * advice_present + b_ec * expert * advice_correct) %>%
    mutate(y_prob = plogis(linpred)) %>%
    mutate(y_bin = rbinom(n = n(), size = 1, prob = y_prob))
  dat
}
```

@@

```{r}
library(tidyverse)
set.seed(1)
dat <- simulate(n_subjects = 500, n_items = 500,
  sd_u0s = 0.5, sd_u0i = 0.5)
dat %>% 
  mutate(condition = fct_cross(
    factor(expert), factor(advice_present), factor(advice_correct))) %>%
  mutate(condition = fct_recode(condition,
    "no expert, no advice" = "0:0:0", "expert, no advice" = "1:0:0", 
    "no expert, wrong advice" = "0:1:0", "expert, wrong advice" = "1:1:0",
    "no expert, correct advice" = "0:1:1", "expert, correct advice" = "1:1:1")) %>% 
  group_by(condition) %>%
  summarize(relative_frequency = sum(y_bin) / n())
```

@@

library(ggdist)
dat %>% 
  mutate(condition = fct_cross(
    factor(expert), factor(advice_present), factor(advice_correct))) %>%
  mutate(condition = fct_recode(condition,
    "no expert, no advice" = "0:0:0", "expert, no advice" = "1:0:0", 
    "no expert, wrong advice" = "0:1:0", "expert, wrong advice" = "1:1:0",
    "no expert, correct advice" = "0:1:1", "expert, correct advice" = "1:1:1")) %>% 
  ggplot(aes(x = y_prob, y = condition)) +
  stat_histinterval(point_interval = "mean_qi", slab_color = "gray45") +
  scale_x_continuous(breaks = seq(0, 1, 0.1), limits = c(0, 1))
```

```{r}
#| label: fig-margdist2
#| fig-cap: "Marginal distributions including means, 66% and 95% confidence intervals for all experimental conditions while setting the standard deviation of item random intercepts to 0.01."
set.seed(1)
dat <- simulate(n_subjects = 500, n_items = 500, sd_u0i = 0.01)
dat %>% 
  mutate(condition = fct_cross(
    factor(expert), factor(advice_present), factor(advice_correct))) %>%
  mutate(condition = fct_recode(condition,
    "no expert, no advice" = "0:0:0", "expert, no advice" = "1:0:0", 
    "no expert, wrong advice" = "0:1:0", "expert, wrong advice" = "1:1:0",
    "no expert, correct advice" = "0:1:1", "expert, correct advice" = "1:1:1")) %>% 
  ggplot(aes(x = y_prob, y = condition)) +
  stat_histinterval(point_interval = "mean_qi", slab_color = "gray45") +
  scale_x_continuous(breaks = seq(0, 1, 0.1), limits = c(0, 1))
```

```{r}
#| include: false
set.seed(1)
dat <- simulate(n_subjects = 500, n_items = 500, sd_u0s = 0.01)
dat %>% 
  mutate(condition = fct_cross(
    factor(expert), factor(advice_present), factor(advice_correct))) %>%
  mutate(condition = fct_recode(condition,
    "no expert, no advice" = "0:0:0", "expert, no advice" = "1:0:0", 
    "no expert, wrong advice" = "0:1:0", "expert, wrong advice" = "1:1:0",
    "no expert, correct advice" = "0:1:1", "expert, correct advice" = "1:1:1")) %>% 
  ggplot(aes(x = y_prob, y = condition)) +
  stat_histinterval(point_interval = "mean_qi", slab_color = "gray45") +
  scale_x_continuous(breaks = seq(0, 1, 0.1), limits = c(0, 1))
```

```{r}
#| label: fig-margdist3
#| echo: false
#| fig-cap: "Marginal distributions including means, 66% and 95% confidence intervals for all experimental conditions while setting the standard deviation of subject and item random intercepts to 3."
set.seed(1)
dat <- simulate(n_subjects = 500, n_items = 500, sd_u0s = 3, sd_u0i = 3)
dat %>% 
  mutate(condition = fct_cross(
    factor(expert), factor(advice_present), factor(advice_correct))) %>%
  mutate(condition = fct_recode(condition,
    "no expert, no advice" = "0:0:0", "expert, no advice" = "1:0:0", 
    "no expert, wrong advice" = "0:1:0", "expert, wrong advice" = "1:1:0",
    "no expert, correct advice" = "0:1:1", "expert, correct advice" = "1:1:1")) %>% 
  ggplot(aes(x = y_prob, y = condition)) +
  stat_histinterval(point_interval = "mean_qi", slab_color = "gray45") +
  scale_x_continuous(breaks = seq(0, 1, 0.1), limits = c(0, 1))
```

@@

```{r}
#| message: false
library(tidyverse)
set.seed(1)
dat <- simulate(n_subjects = 100, n_items = 50)
```

```{r}
#| message: false
library(lme4)
f <- y_bin ~ 1 + expert + advice_present + advice_correct + 
  expert:advice_present + expert:advice_correct +
  (1|subject) + (1|item)
```

```{r}
fit <- glmer(f, data = dat, family = "binomial")
```

```{r}
summary(fit)
```

@@

```{r}
pred_df <- data.frame(advice_present = c(1, 0), advice_correct = c(1, 0), 
  expert = c(1, 1))
pred_df
preds <- predict(fit, newdata = pred_df, type = "response", re.form = NA)
preds
preds[1] - preds[2]
```

```{r}
library(tidyverse)
library(marginaleffects)
pred_df <- expand_grid(advice_present = 0:1, advice_correct = 0:1, expert = 0:1)
pred_df
probs <- predictions(fit, newdata = pred_df, type = "response", re.form = NA)
probs
```

```{r}
probs %>% 
  hypotheses(hypothesis = c(
    "b8 = b2",
    "b2 = b6",
    "b7 = b1",
    "b1 = b5"),
    equivalence = c(0, 0))
```

@@

```{r}
sim_and_analyse <- function(
  formula_chr = "y_bin ~ 1 + expert + advice_present + advice_correct + 
    expert:advice_present + expert:advice_correct + (1|subject) + (1|item)",
  contrasts = c("b8 = b2", "b2 = b6", "b7 = b1", "b1 = b5"), ...){
  require(lme4)
  require(marginaleffects)
  require(tidyr)
  # simulate data
  dat <- simulate(...)
  # fit model
  model <- glmer(as.formula(formula_chr), data = dat, family = "binomial")
  # compute contrasts
  contr_df <- expand_grid(advice_present = 0:1, advice_correct = 0:1,
    expert = 0:1)
  predictions(model, newdata = contr_df, type = "response", re.form = NA) %>%
    hypotheses(hypothesis = contrasts, equivalence = c(0, 0)) %>%
    data.frame()
}
```

@@

```{r}
#| message: false
library(furrr)
sim_result <- crossing(
  rep = 1:5,
  n_subjects = c(100, 150, 200, 250),
  n_items = c(10, 30, 50, 70)
) %>%
  mutate(res = future_pmap(., sim_and_analyse, 
    .options = furrr_options(seed = TRUE))) %>%
  unnest(col = res)
```

@@

```{r}
#| label: fig-finalpwr
#| fig-cap: "Simulation-based power estimates including 95% confidence interval of the case study for different numbers of subjects and items, based on a significance level of 0.05."
library(binom)
alpha <- 0.05
power <- sim_result %>%
  pivot_wider(names_from = term, names_sep = "_", 
    values_from = estimate:p.value.equiv) %>%
  group_by(n_subjects, n_items) %>% 
  summarise(
    power = mean(`p.value.noninf_b1=b5` < alpha & 
        `p.value.noninf_b8=b2` < alpha & `p.value.noninf_b2=b6` < alpha & 
        `p.value.noninf_b7=b1` < alpha), 
    n_sig = sum(`p.value.noninf_b1=b5` < alpha & 
        `p.value.noninf_b8=b2` < alpha & `p.value.noninf_b2=b6` < alpha & 
        `p.value.noninf_b7=b1` < alpha),
    n = n(),
    ci.lwr = binom.confint(n_sig, n, method = "wilson")$lower,
    ci.upr = binom.confint(n_sig, n, method = "wilson")$upper, 
    .groups = "drop")
power %>%
  mutate(across(c(n_subjects, n_items), factor)) %>%
  ggplot(aes(n_subjects, n_items, fill = power)) +
  geom_tile() +
  geom_text(aes(label = sprintf("%.2f \n [%.2f; %.2f]", 
                                power, ci.lwr, ci.upr)), 
    color = "white", size = 4) +
  scale_fill_viridis_c(limits = c(0, 1)) +
  xlab("number of subjects") + ylab("number of items")
```

@@

```{r}
#| label: fig-finalprecision
#| fig-cap: "Simulation-based precision estimates (expected width of confidence intervals) including 95% confidence interval of the case study for different numbers of subjects and items, based on a confidence level of 0.95."
precision <- sim_result %>%
  pivot_wider(names_from = term, names_sep = "_", 
    values_from = estimate:p.value.equiv) %>%
  group_by(n_subjects, n_items) %>%
  mutate(width = `conf.high_b8=b2` - `conf.low_b8=b2`) %>%
  summarise(precision = mean(width),
    ci.lwr = t.test(width)$conf.int[1],
    ci.upr = t.test(width)$conf.int[2], 
    .groups = "drop")
precision %>%
  mutate(across(c(n_subjects, n_items), factor)) %>%
  ggplot(aes(n_subjects, n_items, fill = precision)) +
  geom_tile() +
  geom_text(aes(label = sprintf("%.2f \n [%.2f; %.2f]", 
                                precision, ci.lwr, ci.upr)), 
    color = "white", size = 4) +
  scale_fill_viridis_c(limits = c(0, 0.3), direction = -1) +
  guides(fill = guide_legend(reverse=FALSE))
```

>>>>

---
title: "R code from PRACTICE sections"
execute:
  eval: false
---

::: {.callout-note}
This Notebook contains only the R code from the PRACTICE sections of the manuscript. The number of samples and repetitions for all simulations have been greatly reduced to ensure that the code runs quickly and with limited computational resources.
:::

@@

```{r}
simulate <- function(n_subjects = 100, n_items = 50,
  b_0 = 0.847, b_e = 1.350, b_a = -1.253, b_c = 2.603,
  b_ea = 0.790, b_ec = -1.393,
  sd_u0s = 0.5, sd_u0i = 0.5, ...){
  require(dplyr)
  require(faux)
  # simulate design
  dat <- add_random(subject = n_subjects, item = n_items) |>
    add_between("subject", expert = c(1, 0), .prob = c(0.25, 0.75)) |>
    mutate(advice_present = rbinom(n(), 1, prob = 2/3)) |>
    mutate(advice_correct = if_else(advice_present == 1L, 
                                    rbinom(n(), 1L, prob = 0.8), 0L)) |>
    # add random effects
    add_ranef("subject", u0s = sd_u0s) |>
    add_ranef("item", u0i = sd_u0i) |>
    # compute dependent variable
    mutate(linpred = b_0 + u0i + u0s +
        b_e * expert + b_a * advice_present + b_c * advice_correct +
        b_ea * expert * advice_present + b_ec * expert * advice_correct) |>
    mutate(y_prob = plogis(linpred)) |>
    mutate(y_bin = rbinom(n = n(), size = 1, prob = y_prob))
  dat
}
```

@@

```{r}
library(tidyverse)
set.seed(1)
dat <- simulate(n_subjects = 500, n_items = 500,
  sd_u0s = 0.5, sd_u0i = 0.5)
dat |> 
  mutate(condition = fct_cross(
    factor(expert), factor(advice_present), factor(advice_correct))) |>
  mutate(condition = fct_recode(condition,
 "no expert, no advice" = "0:0:0", "expert, no advice" = "1:0:0", 
 "no expert, wrong advice" = "0:1:0", "expert, wrong advice" = "1:1:0",
 "no expert, correct advice" = "0:1:1", "expert, correct advice" = "1:1:1")) |> 
  group_by(condition) |>
  summarize(relative_frequency = sum(y_bin) / n())
```

@@

library(ggdist)
dat |> 
  mutate(condition = fct_cross(
    factor(expert), factor(advice_present), factor(advice_correct))) |>
  mutate(condition = fct_recode(condition,
 "no expert, no advice" = "0:0:0", "expert, no advice" = "1:0:0", 
 "no expert, wrong advice" = "0:1:0", "expert, wrong advice" = "1:1:0",
 "no expert, correct advice" = "0:1:1", "expert, correct advice" = "1:1:1")) |> 
  ggplot(aes(x = y_prob, y = condition)) +
  stat_histinterval(point_interval = "mean_qi", slab_color = "gray45",
    breaks = "Sturges") +
  scale_x_continuous(breaks = seq(0, 1, 0.1), limits = c(0, 1))
```

```{r}
#| label: fig-margdist2
#| fig-cap: "Marginal distributions including means, 66% and 95% confidence intervals for all experimental conditions while setting the standard deviation of item random intercepts to 0.01."
set.seed(1)
dat <- simulate(n_subjects = 500, n_items = 500, sd_u0i = 0.01)
dat |> 
  mutate(condition = fct_cross(
    factor(expert), factor(advice_present), factor(advice_correct))) |>
  mutate(condition = fct_recode(condition,
    "no expert, no advice" = "0:0:0", "expert, no advice" = "1:0:0", 
    "no expert, wrong advice" = "0:1:0", "expert, wrong advice" = "1:1:0",
    "no expert, correct advice" = "0:1:1", "expert, correct advice" = "1:1:1")) |> 
  ggplot(aes(x = y_prob, y = condition)) +
  stat_histinterval(point_interval = "mean_qi", slab_color = "gray45",
    breaks = "Sturges") +
  scale_x_continuous(breaks = seq(0, 1, 0.1), limits = c(0, 1))
```

```{r}
#| include: false
set.seed(1)
dat <- simulate(n_subjects = 500, n_items = 500, sd_u0s = 0.01)
dat |> 
  mutate(condition = fct_cross(
    factor(expert), factor(advice_present), factor(advice_correct))) |>
  mutate(condition = fct_recode(condition,
    "no expert, no advice" = "0:0:0", "expert, no advice" = "1:0:0", 
    "no expert, wrong advice" = "0:1:0", "expert, wrong advice" = "1:1:0",
    "no expert, correct advice" = "0:1:1", "expert, correct advice" = "1:1:1")) |> 
  ggplot(aes(x = y_prob, y = condition)) +
  stat_histinterval(point_interval = "mean_qi", slab_color = "gray45",
    breaks = "Sturges") +
  scale_x_continuous(breaks = seq(0, 1, 0.1), limits = c(0, 1))
```

```{r}
#| label: fig-margdist3
#| echo: false
#| fig-cap: "Marginal distributions including means, 66% and 95% confidence intervals for all experimental conditions while setting the standard deviation of subject and item random intercepts to 3."
set.seed(1)
dat <- simulate(n_subjects = 500, n_items = 500, sd_u0s = 3, sd_u0i = 3)
dat |> 
  mutate(condition = fct_cross(
    factor(expert), factor(advice_present), factor(advice_correct))) |>
  mutate(condition = fct_recode(condition,
    "no expert, no advice" = "0:0:0", "expert, no advice" = "1:0:0", 
    "no expert, wrong advice" = "0:1:0", "expert, wrong advice" = "1:1:0",
    "no expert, correct advice" = "0:1:1", "expert, correct advice" = "1:1:1")) |> 
  ggplot(aes(x = y_prob, y = condition)) +
  stat_histinterval(point_interval = "mean_qi", slab_color = "gray45",
    breaks = "Sturges") +
  scale_x_continuous(breaks = seq(0, 1, 0.1), limits = c(0, 1))
```

@@

```{r}
#| message: false
library(tidyverse)
library(lme4)
set.seed(1)
dat <- simulate(n_subjects = 100, n_items = 50)
f <- y_bin ~ 1 + expert + advice_present + advice_correct + 
  expert:advice_present + expert:advice_correct +
  (1|subject) + (1|item)
fit <- glmer(f, data = dat, family = "binomial")
summary(fit)
```

@@

```{r}
grid1 <- data.frame(advice_present = c(1, 0), advice_correct = c(1, 0), 
  expert = c(1, 1))
grid1
pred <- predict(fit, newdata = grid1, type = "response", re.form = NA)
pred
pred[1] - pred[2]
```

```{r}
grid2 <- expand_grid(advice_present = 0:1, 
  advice_correct = 0:1, expert = 0:1)
grid2
preds <- predictions(fit, newdata = grid2, 
  type = "response", re.form = NA)
as.data.frame(preds)[,c(2,7:11)]
```

```{r}
contrasts <- preds |> 
  hypotheses(hypothesis = c(
    "b8 = b2",  # (correct advice, expert) - (no advice, expert)
    "b2 = b6",  # (no advice, expert) - (wrong advice, expert) 
    "b7 = b1",  # (correct advice, no expert) - (no advice, no expert)
    "b1 = b5"), # (no advice, no expert) - (wrong advice, no expert)
    equivalence = c(0, 0))
contrasts
```

@@

```{r}
sim_and_analyse <- function(
  formula_chr = "y_bin ~ 1 + expert + advice_present + advice_correct + 
    expert:advice_present + expert:advice_correct + (1|subject) + (1|item)",
  contrasts = c("b8 = b2", "b2 = b6", "b7 = b1", "b1 = b5"), ...){
  require(lme4)
  require(marginaleffects)
  require(tidyr)
  # simulate data
  dat <- simulate(...)
  # fit model
  model <- glmer(as.formula(formula_chr), data = dat, family = "binomial")
  # compute contrasts
  contr_df <- expand_grid(advice_present = 0:1, advice_correct = 0:1,
    expert = 0:1)
  predictions(model, newdata = contr_df, type = "response", re.form = NA) |>
    hypotheses(hypothesis = contrasts, equivalence = c(0, 0)) |>
    data.frame()
}
```

@@

```{r}
#| message: false
library(furrr)
sim_result <- crossing(
  rep = 1:5,
  n_subjects = c(100, 150, 200, 250),
  n_items = c(10, 30, 50, 70)
) |>
  mutate(res = future_pmap(., sim_and_analyse, 
    .options = furrr_options(seed = TRUE))) |>
  unnest(col = res)
```

@@

```{r}
#| label: fig-finalpwr
#| fig-cap: "Simulation-based power estimates including 95% confidence interval of the case study for different numbers of subjects and items, based on a significance level of 0.05."
library(binom)
alpha <- 0.05
power <- sim_result |>
  pivot_wider(names_from = term, names_sep = "_", 
    values_from = estimate:p.value.equiv) |>
  group_by(n_subjects, n_items) |> 
  summarise(
    power = mean(`p.value.noninf_b1=b5` < alpha & 
        `p.value.noninf_b8=b2` < alpha & `p.value.noninf_b2=b6` < alpha & 
        `p.value.noninf_b7=b1` < alpha), 
    n_sig = sum(`p.value.noninf_b1=b5` < alpha & 
        `p.value.noninf_b8=b2` < alpha & `p.value.noninf_b2=b6` < alpha & 
        `p.value.noninf_b7=b1` < alpha),
    n = n(),
    ci.lwr = binom.confint(n_sig, n, method = "wilson")$lower,
    ci.upr = binom.confint(n_sig, n, method = "wilson")$upper, 
    .groups = "drop")
power |>
  mutate(across(c(n_subjects, n_items), factor)) |>
  ggplot(aes(n_subjects, n_items, fill = power)) +
  geom_tile() +
  geom_text(aes(label = sprintf("%.2f \n [%.2f; %.2f]", 
                                power, ci.lwr, ci.upr)), 
    color = "white", size = 4) +
  scale_fill_viridis_c(limits = c(0, 1)) +
  xlab("number of subjects") + ylab("number of items")
```

@@

```{r}
#| label: fig-finalprecision
#| fig-cap: "Simulation-based precision estimates (expected width of confidence intervals) including 95% confidence interval of the case study for different numbers of subjects and items, based on a confidence level of 0.95."
precision <- sim_result |>
  pivot_wider(names_from = term, names_sep = "_", 
    values_from = estimate:p.value.equiv) |>
  group_by(n_subjects, n_items) |>
  mutate(width = `conf.high_b8=b2` - `conf.low_b8=b2`) |>
  summarise(precision = mean(width),
    ci.lwr = t.test(width)$conf.int[1],
    ci.upr = t.test(width)$conf.int[2], 
    .groups = "drop")
precision |>
  mutate(across(c(n_subjects, n_items), factor)) |>
  ggplot(aes(n_subjects, n_items, fill = precision)) +
  geom_tile() +
  geom_text(aes(label = sprintf("%.2f \n [%.2f; %.2f]", 
                                precision, ci.lwr, ci.upr)), 
    color = "white", size = 4) +
  scale_fill_viridis_c(limits = c(0, 0.3), direction = -1) +
  guides(fill = guide_legend(reverse=FALSE))
```
